\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-23 papers/jss/article.Rnw
mix of minor fixes from the train ride
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@826 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Dirk Eddelbuettel \\
  Debian Project \\
  River Forest, IL, USA\\
  E-mail: \email{edd@debian.org}\\
  URL: \url{http://dirk.eddelbuettel.com}\\
  \\
  Murray Stokely\\
  Google, Inc.\\
  1600 Amphitheatre Parkway\\
  Mountain View, CA 94040\\
  USA\\
  E-mail: \email{mstokely@google.com}\\
  URL: \url{http://www.stokely.org/}\\
  \\
  Jeroen Ooms\\
  UCLA Department of Statistics\\
  University of California\\
  E-mail: \email{jeroen.ooms@stat.ucla.edu}\\
  URL: \url{http://jeroenooms.github.io}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

@@

This paper describes an \proglang{R} interface to Protocol Buffers,
and is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive \proglang{R} interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between \proglang{R} and other languages.  Section~\ref{sec:evaluation} introduces a
general \proglang{R} language schema for serializing arbitrary \proglang{R} objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:mapreduce}
and \ref{sec:opencpu} provide real-world use cases of \CRANpkg{RProtoBuf}
in MapReduce and web service environments, respectively, before
Section~\ref{sec:summary} concludes.

@@

\Sexpr{m} data sets can be converted to Protocol Buffers
(\Sexpr{format(100*m/n,digits=1)}\%) without loss of information. Upon closer
inspection, all other datasets are objects of class \texttt{nfnGroupedData}.
This class represents a special type of data frame that has some additional 
attributes used by the \pkg{nlme} package, among which a \emph{formula} object.
Because formulas are \proglang{R} \emph{language} objects, they have little meaning to
other systems, and are not supported by the \texttt{rexp.proto} descriptor.
When \texttt{serialize\_pb} is used on objects of this class, it will serialize
the data frame and all attributes, except for the formula.

@@

Table~\ref{tab:compression} shows the sizes of 50 sample \proglang{R} datasets as
returned by object.size() compared to the serialized sizes.
%The summary compression sizes are listed below, and a full table for a
%sample of 50 datasets is included on the next page.  
Sizes are comparable but Protocol Buffers provide simple getters and setters
in multiple languages instead of requiring other programs to parse the R
serialization format. % \citep{serialization}.
One takeaway from this table is that the universal \proglang{R} object schema
included in RProtoBuf does not in general provide
any significant saving in file size compared to the normal serialization
mechanism in R.
% redundant: which is seen as equally compact.
The benefits of RProtoBuf accrue more naturally in applications where
multiple programming languages are involved, or when a more concise
application-specific schema has been defined.  The example in the next
section provides both of these conditions.

@@

<<echo=FALSE,fig=TRUE,width=8,height=4>>=
require(RProtoBuf)
require(HistogramTools)
readProtoFiles(package="HistogramTools")
hist <- HistogramTools.HistogramState$read("hist.pb")
plot(as.histogram(hist))
@

@@

As described earlier, the primary application of Protocol Buffers is data
interchange in the context of inter-system communications.  Network protocols
such as HTTP provide mechanisms for client-server communication, i.e. how to
initiate requests, authenticate, send messages, etc.  However, network
protocols generally do not regulate the \emph{content} of messages: they
allow transfer of any media type, such as web pages, static files or
multimedia content.  When designing systems where various components require
exchange of specific data structures, we need something on top of the network
protocol that prescribes how these structures are to be represented in
messages (buffers) on the network. Protocol Buffers solve exactly this
problem by providing a cross-platform method for serializing arbitrary
structures into well defined messages, which can then be exchanged using any
protocol. The descriptors (\texttt{.proto} files) are used to formally define
the interface of a remote API or network application. Libraries to parse and
generate protobuf messages are available for many programming languages,
making it relatively straightforward to implement clients and servers.

@@

%\section[About Java]{About \proglang{Java}}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

>>>>

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Dirk Eddelbuettel \\
  Debian Project \\
  River Forest, IL, USA\\
  E-mail: \email{edd@debian.org}\\
  URL: \url{http://dirk.eddelbuettel.com}\\
  \\
  Murray Stokely\\
  Google, Inc.\\
  1600 Amphitheatre Parkway\\
  Mountain View, CA, USA\\
  E-mail: \email{mstokely@google.com}\\
  URL: \url{http://www.stokely.org/}\\
  \\
  Jeroen Ooms\\
  UCLA Department of Statistics\\
  University of California\\
  Los Angeles, CA, USA\\
  E-mail: \email{jeroen.ooms@stat.ucla.edu}\\
  URL: \url{http://jeroenooms.github.io}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

@@

This paper describes an \proglang{R} interface to Protocol Buffers,
and is organized as follows. Section~\ref{sec:protobuf}
provides a general high-level overview of Protocol Buffers as well as a basic
motivation for their use.
Section~\ref{sec:rprotobuf-basic} describes the interactive \proglang{R} interface
provided by the \CRANpkg{RProtoBuf} package, and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
details the implementation details of the main S4 classes and methods
contained in this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between \proglang{R} and other languages.  Section~\ref{sec:evaluation} introduces a
general \proglang{R} language schema for serializing arbitrary \proglang{R} objects and evaluates
it against the serialization capbilities built directly into R.  Sections~\ref{sec:mapreduce}
and \ref{sec:opencpu} provide real-world use cases of \CRANpkg{RProtoBuf}
in MapReduce and web service environments, respectively, before
Section~\ref{sec:summary} concludes.

@@

\Sexpr{m} data sets can be converted to Protocol Buffers
(\Sexpr{format(100*m/n,digits=1)}\%) without loss of information. Upon closer
inspection, all other datasets are objects of class \texttt{nfnGroupedData}.
This class represents a special type of data frame that has some additional 
attributes (such as a \emph{formula} object) used by the \pkg{nlme} package.
Because formulas are \proglang{R} \emph{language} objects, they have little meaning to
other systems, and are not supported by the \texttt{rexp.proto} descriptor.
When \texttt{serialize\_pb} is used on objects of this class, it will serialize
the data frame and all attributes, except for the formula.

@@

Table~\ref{tab:compression} shows the sizes of 50 sample \proglang{R} datasets as
returned by object.size() compared to the serialized sizes.
%The summary compression sizes are listed below, and a full table for a
%sample of 50 datasets is included on the next page.  
Sizes are comparable but Protocol Buffers provide simple getters and setters
in multiple languages instead of requiring other programs to parse the R
serialization format. % \citep{serialization}.
One takeaway from this table is that the universal \proglang{R} object schema
included in \pkg{RProtoBuf} does not in general provide
any significant saving in file size compared to the normal serialization
mechanism in R.
% redundant: which is seen as equally compact.
The benefits of \pkg{RProtoBuf} accrue more naturally in applications where
multiple programming languages are involved, or when a more concise
application-specific schema has been defined.  The example in the next
section satisfies both of these conditions.

@@

\begin{center}
<<echo=FALSE,fig=TRUE,width=8,height=4>>=
require(RProtoBuf)
require(HistogramTools)
readProtoFiles(package="HistogramTools")
hist <- HistogramTools.HistogramState$read("hist.pb")
plot(as.histogram(hist))
@
\end{center}

@@

As described earlier, the primary application of Protocol Buffers is data
interchange in the context of inter-system communications.  Network protocols
such as HTTP provide mechanisms for client-server communication, i.e.~how to
initiate requests, authenticate, send messages, etc.  However, network
protocols generally do not regulate the \emph{content} of messages: they
allow transfer of any media type, such as web pages, static files or
multimedia content.  When designing systems where various components require
exchange of specific data structures, we need something on top of the network
protocol that prescribes how these structures are to be represented in
messages (buffers) on the network. Protocol Buffers solve exactly this
problem by providing a cross-platform method for serializing arbitrary
structures into well defined messages, which can then be exchanged using any
protocol. The descriptors (\texttt{.proto} files) are used to formally define
the interface of a remote API or network application. Libraries to parse and
generate protobuf messages are available for many programming languages,
making it relatively straightforward to implement clients and servers.

@@
